Application of a Cut-Cell Immersed Boundary Method for Wildfire Simulation over Complex Terrain

### 2020-12-08 Response_to_A_Maranghides.tex

\documentclass[12pt]{article}

\setlength{\textwidth}{6.25in}
\setlength{\textheight}{8.75in}
\setlength{\topmargin}{0.in}
\setlength{\headheight}{0.in}
\setlength{\headsep}{0.in}
\setlength{\parindent}{0.in}
\setlength{\oddsidemargin}{0.0in}
\setlength{\evensidemargin}{0.0in}
\setlength{\parskip}{\baselineskip}
\setlength{\leftmargini}{\parindent} % Controls the indenting of the "bullets" in a list

\usepackage{xcolor}
\newcommand\hl[1]{\textcolor{red}{#1}}

\pagestyle{empty}

\begin{document}

\begin{center}
{\bf Response to comments by WERB Reviewer A. Maranghides}
\end{center}

I focused my review on the later part of the paper, because of my technical expertise and also because of context/ impact.

I am specifically interested in the case of the Cogoleto Fire. I think the case presented here is \textbf{oversimplified and can potentially direct the reader to make certain assumptions about the model capabilities}. Even though this is a small part of the paper in many ways it is maybe the most important in terms of representing the model usefulness.

What is lacking here are:
%
\begin{itemize}

   \item Information about the scale of the event: how large was the fire from origin to furthest point, how long it took to reach the final perimeter. A scale on the figure 4 would be helpful.

   From origin to furthest point measured distance is ~ 1200 m.

   \item Wind context: was the single measurement consistent both in speed and direction throughout the event. How representative was this single point measurement, given the second closest weather station and how that data compared. was the station at a specific location (ravine or local acceparation zone). I can tell you that in CA, in the Camp Fire (as an example here) 5 -8 km can make all the difference in the world (100 km/h vs 10 km/h)

   \item information on defensive actions. Defensive actions not only impact the final perimeter but can also impact the direction as firefighters typically try to direct the fire away from high value targets. This can be done with both ground and aerial resources. I do not know what was used and when.

   \item There is no time scale provided on the runs or event progression so it is very difficult to judge what is presented.

   \item There is no mention of spotting within the fire perimeter and also outside. Short, intermediate and far field spotting can play critical roles in the evolution of a wildfire and/or WUI incident and in my mind is one of the key issues that realistically limits the applicability of predictive tools on that scale. Since we cannot resolve where embers will result in ignitions and the location of the actual ignitions can have dramatic impact on how the event develops, not knowing limits the usefulness of these tools. A probabilistic approach does not improve the usefulness in my mind.

   \item Explicit listing of the model limitations of dealing with fuel breaks both natural and manmade. While I appreciate that if the wind was truly 100km/h fuel breaks may have had little impact on the overall fire spread, the effect on fuel breaks can still provide critical information to understanding how fire behaved and should be discussed.

   \item I would also like to see some statement explicitly addressing the complexity of dealing with non-uniform fuels (not only vegetative but also WUI). Even using the level set approach one needs to be mindful of the potential change in fire spread as the fire migrates between different fuels.

\end{itemize}

My review recommendations are the following:

\begin{enumerate}
    \item[A:] Explicitly address the above mentioned issues.

    \item[B:] Use this opportunity to highlight the value of developing these large scale outdoor tools which are (in my mind):
    \item[B1.] Adding value to prescribed burns planning, where conditions are more benign (specifically wind) and data can be collected to refine and develop the models
    \item[B2.] Use the ability to model complex geometries at a multi-parcel level using physics to highly resolve fire spread and ultimately ignition.

\hl{For the previous two questions Kevin added a paragraph: "The strategy that has been adopted in FDS is to accommodate simulation options ranging from a level set calculation of fire spread over tens of kilometers that can be run in minutes, similar to models like BEHAVE developed by the U.S. Forest Service, all the way to detailed simulations of fire spread at sub-meter resolution for relatively short time periods (minutes to hours) and small areas (tens of hectares)." at the end of the Cogoleto section}

\end{enumerate}

Even if you are not interested/agree to go with B, at a minimum consider addressing A. While it may appear that I have been critical, I think the capabilities added to deal with complex geometries can add very much needed value to FDS at a different scale where we can test and reliably validate the code.

\end{document}

### 2020-12-08 Vanella_Article_Atmosphere_2020.tex
Update on Overleaf.

@@

\abstract{A method for the large-eddy simulation (LES) of wildfire spread over complex terrain is presented. In this scheme, a cut-cell immersed boundary method (CC-IBM) is used to render the complex terrain, defined by a tessellation, on a rectilinear Cartesian grid. Discretization of scalar transport equations for chemical species is done via a finite volume scheme on cut-cells defined by the intersection of the terrain geometry and the Cartesian cells. Momentum transport and heat transfer close to the immersed terrain are handled using dynamic wall models and a direct forcing immersed boundary method. Further, atmospheric wind conditions are specified using a mean forcing concept. Additionally, three basic approaches have been explored to model fire spread: (1) representing the vegetation as a collection of Lagrangian particles, (2) representing the vegetation as a semi-porous boundary, and (3) representing the fire spread using a level set method in which the fire spreads as a function of terrain slope, vegetation type, and wind speed. Several test and validation cases are reported to demonstrate the capabilities of this novel wildfire simulation methodology.}

@@

\newcommand{\dV}{{\rm d}V}
\newcommand{\dS}{{\rm d}S}

@@

In the past few decades, wildland and wildland-urban interface (WUI) fires have received substantial attention due to their destructive nature and extensive cost in lives and property~\cite{thomas_2017,mcdermott_2019,richards_2020}. A large body of literature can be found on the subject, as seen in various review articles~\cite{Papadopoulos_2011,Bakhshaii_2019,mcdermott_2019}. Due to the complex nature of fire, limited computing resources, and the needs of planners and first responders, most models of wildfires have historically relied on simplified field-tested rules and correlations. Among these, the rate of spread model of Rothermel~\cite{Rothermel:1972} assumes quasi-steady state surface fire conditions and takes as inputs fuel properties, wind, terrain slope, and moisture that can be measured {\em in situ}. Fuel models have been cataloged~\cite{Anderson:1982}, and their parameters defined so that model users can select and combine the most appropriate listed fuels. Rothermel's rate of spread formula is still widely used, and has been implemented in several simulation tools~\cite{Finney:FARSITE,Bova:IJWF2015,mcgratta_2013,Coen:2,Coen:2015,LAUTENBERGER_2013,Coen:2013,Mandel:2009,Mandel:2011,Mandel:2014,Kochanski:2016}, enabling a range of planning and operational forecasting capabilities.

Wildland fire modeling typically focuses on tracking the fire front, which can have a complex shape that moves and deforms based on local conditions like fuel loading, moisture content, wind, and terrain. The fire front requires a mathematical representation. The Lagrangian approach is based on a polygonal mesh of control markers that represents the front~\cite{Finney:FARSITE,Bova:IJWF2015}. On the other hand, an Eulerian representation allows for a fixed two-dimensional mesh on which the fire front is defined implicitly by a scalar field. This scalar field is called the level set function and numerical methods that compute the function are called Level Set Methods (LSM)~\cite{Sethian:1999,Osher:2006}. Some wildfire solvers that use LSM to track the fire front are described in references~\cite{coen_2013,Bova:IJWF2015,mcgratta_2013,LAUTENBERGER_2013}. A practical comparison of the two techniques as implemented in the Fire Dynamics Simulator (FDS)~\cite{Mell:IJWF2007} and FARSITE~\cite{Finney:FARSITE} is provided in reference~\cite{Bova:IJWF2015}.

@@

@@

@@

Consider a mixture composed of $N$ chemical species $\alpha$, moving on a fixed point $\mathbf{x}$ in space with a mass weighted average velocity $\mathbf{u}(\mathbf{x},t)$. If $\rho(\mathbf{x},t)$ is the mixture density, the mass fraction for species $\alpha$ is $Y_\alpha = \rho_\alpha / \rho$, where $\rho_\alpha(\mathbf{x},t)$ is the species mass density and $\rho = \sum \rho_\alpha$. The scalar transport and momentum equations take the form:
\begin{eqnarray}
   \frac{\partial \rho Y_\alpha}{ \partial t} + \nabla \cdot ( \rho Y_\alpha  \mathbf{u} ) &=& - \nabla \cdot \mathbf{J_{d}}_\alpha + \dot{m}_\alpha'''  +
    \dot{m}_{b,\alpha}'''\; , \; \alpha=1,\dots,N \label{eqn:spectran} \\
    \frac{\partial \mathbf{u}}{\partial t} - \mathbf{u} \times \boldsymbol{\omega} + \nabla H - \tilde{p} \, \nabla \left( 1/\rho\right) &=&
    \frac{1}{\rho} \left[ (\rho-\rho_0) \mathbf{g} + \mathbf{f}_{b} + \nabla \cdot \boldsymbol{\tau} \right] \label{eqn:momtran}
\end{eqnarray}
where $\mathbf{J_{d}}_\alpha=- \rho D_\alpha \boldsymbol{\nabla} Y_\alpha$ is the diffusive flux for component $\alpha$. Fick's Law for binary diffusion with respect to a background species is assumed, and $D_\alpha$ is the diffusivity of $\alpha$ with respect to the background species.  To enforce realizability, $\sum_\alpha \mathbf{J_{d}}_\alpha = 0$, errors in diffusive transport are lumped into the most abundant species locally~\cite{McDermott:2015}. The volumetric combustion source term for species $\alpha$ is $\dot{m}_\alpha'''(\mathbf{x},t)$  and $\dot{m}_{b,\alpha}'''$ is the contribution to species $\alpha$ from subgrid particle gasification. In the momentum equation the term $H=|\mathbf{u}|^2/2 + \tilde{p}/\rho$, $\tilde{p}$ is the perturbation pressure, and $\rho_0(z)$ is an ambient density of the atmosphere.  The gravity vector is $\mathbf{g}=(0,0,g_z)$, $\mathbf{f}_{b}$ is the particle force term, and $\boldsymbol{\tau}$ is a tensor accounting for molecular and subgrid turbulent stresses parameterized by an effective eddy viscosity model. This last model is employed by default in LES of outdoor fires. Details of the model equations and their derivation are found in Ref.~\cite{mcgratta_2013tr}.

Conservation of energy is achieved by forcing the flow field to obey the following thermodynamic divergence constraint:
\begin{eqnarray}
    ( \nabla \cdot \mathbf{u} )^{\rm th} &=&
    \left[ \frac{1}{\rho c_p T} - \frac{1}{\bar{p}} \right]
    \frac{\partial \bar{p}}{\partial t} + \frac{w \rho_0 g_z}{\rho c_p T} \nonumber \\
    &+& \frac{1}{\rho c_p T} \left[ \dot{q}''' + \dot{q}_b''' - \nabla \cdot \dot{\mathbf{q}}'' - \mathbf{u} \cdot \nabla (\rho h_s) \right] \nonumber \\
    &+& \frac{1}{\rho} \sum_\alpha \left( \frac{\overline{W}}{W_\alpha} - \frac{h_{s,\alpha}}{c_p T} \right) \left[ \dot{m}_\alpha'''  + \dot{m}_{b,\alpha}''' - \nabla \cdot (\mathbf{J_{d}}_\alpha) - \mathbf{u} \cdot \nabla (\rho Y_\alpha) \right] \label{eq:divth}
\end{eqnarray}
where $h_s$ is the local sensible enthalpy, $\dot{q}''', \dot{q}_b'''$ are heat release rates due to combustion and particle gasification, $\dot{\mathbf{q}}''$ is the heat flux vector (conduction, convection and radiation), and $T$ is the local gas temperature. The mixture specific heat at constant pressure and molecular weight are $c_p=\sum_{\alpha =1}^N{c_{p,\alpha} Y_\alpha}$ and $\overline{W}=\left(\sum_{\alpha =1}^N{Y_\alpha /W_\alpha} \right)^{-1}$ respectively. This thermodynamic divergence is derived from and acts as a proxy for the sensible enthalpy evolution equation~\cite{mcdermo_2014}. Equation~(\ref{eq:divth}) is derived by factoring the divergence from the sensible enthalpy equation and applying the ideal gas law.  Given a background pressure and a local mass density, the local temperature is computed from the ideal gas law as $T = \bar{p} \overline{W} / (\rho R) $.

\section{Terrain description and discretization} \label{sec:terraindisc}

>>>>

\usepackage[export]{adjustbox}

@@

\abstract{A method for the large-eddy simulation (LES) of wildfire spread over complex terrain is presented. In this scheme, a cut-cell immersed boundary method (CC-IBM) is used to render the complex terrain, defined by a tessellation, on a rectilinear Cartesian grid. Discretization of scalar transport equations for chemical species is done via a finite volume scheme on cut-cells defined by the intersection of the terrain geometry and the Cartesian cells. Momentum transport and heat transfer close to the immersed terrain are handled using dynamic wall models and a direct forcing immersed boundary method. A new ``open'' convective inflow/outflow method for specifying atmospheric wind boundary conditions is presented. Additionally, three basic approaches have been explored to model fire spread: (1) representing the vegetation as a collection of Lagrangian particles, (2) representing the vegetation as a semi-porous boundary, and (3) representing the fire spread using a level set method in which the fire spreads as a function of terrain slope, vegetation type, and wind speed. Several test and validation cases are reported to demonstrate the capabilities of this novel wildfire simulation methodology.}

@@

\newcommand{\dV}{{\rm d}V}
\newcommand{\dS}{{\rm d}S}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bs}{\mathbf{s}}
\renewcommand{\C}{\mathrm{C}}
\renewcommand{\H}{\mathrm{H}}
\renewcommand{\O}{\mathrm{O}}

@@

In the past few decades, wildland and wildland-urban interface (WUI) fires have received substantial attention due to their destructive nature and extensive cost in lives and property~\cite{thomas_2017,mcdermott_2019,richards_2020}. A large body of literature can be found on the subject, as seen in various review articles~\cite{Papadopoulos_2011,Bakhshaii_2019,mcdermott_2019}. Due to the complex nature of fire, limited computing resources, and the needs of planners and first responders, most models of wildfires have historically relied on simplified field-tested rules and correlations. Among these, the rate of spread model of Rothermel~\cite{Rothermel:1972} assumes quasi-steady state surface fire conditions and takes as inputs fuel properties, wind, terrain slope, and moisture that can be measured {\em in situ}. Fuel models have been cataloged~\cite{Anderson:1982}, and their parameters defined so that model users can select and combine the most appropriate listed fuels. Rothermel's rate of spread formula is still widely used, and has been implemented in several simulation tools~\cite{Finney:FARSITE,Bova:IJWF2015,FDS_Users_Guide,Coen:2,Coen:2015,LAUTENBERGER_2013,Coen:2013,Mandel:2009,Mandel:2011,Mandel:2014,Kochanski:2016}, enabling a range of planning and operational forecasting capabilities.

Wildland fire modeling typically focuses on tracking the fire front, which can have a complex shape that moves and deforms based on local conditions like fuel loading, moisture content, wind, and terrain. The fire front requires a mathematical representation. The Lagrangian approach is based on a polygonal mesh of control markers that represents the front~\cite{Finney:FARSITE,Bova:IJWF2015}. On the other hand, an Eulerian representation allows for a fixed two-dimensional mesh on which the fire front is defined implicitly by a scalar field. This scalar field is called the level set function and numerical methods that compute the function are called Level Set Methods (LSM)~\cite{Sethian:1999,Osher:2006}. Some wildfire solvers that use LSM to track the fire front are described in references~\cite{coen_2013,Bova:IJWF2015,FDS_Users_Guide,LAUTENBERGER_2013}. A practical comparison of the two techniques as implemented in the Fire Dynamics Simulator (FDS)~\cite{Mell:IJWF2007} and FARSITE~\cite{Finney:FARSITE} is provided in reference~\cite{Bova:IJWF2015}.

@@

A number of other approaches to modeling fire spread over complex terrain exist. A recent review, with an emphasis on operational applications, is given in~\cite{Arca_2019}. A range of approximations to modeling both the wind and fire have been employed. The uniqueness of the approach presented here is that, within a single computational tool, a range of approximations to the wind and fire physics can be employed. This can support a direct comparison between models of varying physical fidelity. For example, one can investigate the difference in predicted fire behavior between a simulation that models fire spread and heat release via the explicit accounting of the thermal degradation of vegetation and one that approximates fire spread via an empirically based model.

@@

\subsection{Governing Equations} \label{sec:goveqns}

@@

Consider a mixture composed of $N$ chemical species $\alpha$, moving on a fixed point $\mathbf{x}$ in space with a mass weighted average velocity $\mathbf{u}(\mathbf{x},t)$. If $\rho(\mathbf{x},t)$ is the mixture density, the mass fraction for species $\alpha$ is $Y_\alpha = \rho_\alpha / \rho$, where $\rho_\alpha(\mathbf{x},t)$ is the species mass density and $\rho = \sum \rho_\alpha$. The scalar transport and momentum equations take the form:
\begin{eqnarray}
   \frac{\partial \rho Y_\alpha}{ \partial t} + \nabla \cdot ( \rho Y_\alpha  \mathbf{u} ) &=& - \nabla \cdot \mathbf{J_{d}}_\alpha + \dot{m}_\alpha'''  +
    \dot{m}_{b,\alpha}'''\; , \; \alpha=1,\dots,N \label{eqn:spectran} \\
    \frac{\partial \mathbf{u}}{\partial t} - \mathbf{u} \times \boldsymbol{\omega} + \nabla H - \tilde{p} \, \nabla \left( 1/\rho\right) &=&
    \frac{1}{\rho} \left[ (\rho-\rho_0) \mathbf{g} + \mathbf{f}_{b} + \nabla \cdot \boldsymbol{\tau}^{dev} \right] \label{eqn:momtran}
\end{eqnarray}
where $\mathbf{J_{d}}_\alpha=- \rho D_\alpha \boldsymbol{\nabla} Y_\alpha$ is the diffusive flux for component $\alpha$. Fick's Law for binary diffusion with respect to a background species is assumed, and $D_\alpha$ is the diffusivity of $\alpha$ with respect to the background species.  Mass fractions $Y_\alpha$, solution of equations~\eqref{eqn:spectran}, must obey realizability constraints $0<Y_\alpha<1$ and $\sum_\alpha Y_\alpha=1$. In equations~\eqref{eqn:spectran} starting from a realizable solution, realizability implies $\sum_\alpha \mathbf{J_{d}}_\alpha = 0$. Therefore, to enforce realizability, errors in diffusive transport are lumped into the most abundant species locally~\cite{McDermott:2015}. The volumetric combustion source term for species $\alpha$ is $\dot{m}_\alpha'''(\mathbf{x},t)$  and $\dot{m}_{b,\alpha}'''$ is the contribution to species $\alpha$ from subgrid particle gasification. These correspond to the amount of mass per unit volume and time of species $\alpha$ being added or substracted in a given point $\mathbf{x}$ due to chemical reaction or gasification of solid particles (or droplets), respectively.  In the momentum equation the term $H=|\mathbf{u}|^2/2 + \tilde{p}/\rho$, $\tilde{p}$ is the perturbation pressure, and $\rho_0(z)$ is the modeled height dependent background density of the atmosphere.  Also, $\boldsymbol{\omega}$ refers to the vorticity field. The gravity vector is $\mathbf{g}=(0,0,g_z)$, the buoyancy force term is $ (\rho-\rho_0) \mathbf{g}$, $\mathbf{f}_{b}$ is the term contributed by modeled particle drag forces, and $\boldsymbol{\tau}^{dev}$ is the deviatoric stress tensor accounting for molecular and subgrid turbulent stresses parameterized by an effective eddy viscosity model, discussed below in Sec.~\ref{sec:subgrid}.

Conservation of energy is achieved by forcing the flow field to obey the following thermodynamic divergence constraint:
\begin{eqnarray}
    ( \nabla \cdot \mathbf{u} )^{\rm th} &=&
    \left[ \frac{1}{\rho c_p T} - \frac{1}{\bar{p}} \right]
    \frac{\partial \bar{p}}{\partial t} + \frac{w \rho_0 g_z}{\rho c_p T} \nonumber \\
    &+& \frac{1}{\rho c_p T} \left[ \dot{q}''' + \dot{q}_b''' - \nabla \cdot \dot{\mathbf{q}}'' - \nabla \cdot \dot{\mathbf{q}}_{\rm r}'' - \mathbf{u} \cdot \nabla (\rho h_s) \right] \nonumber \\
    &+& \frac{1}{\rho} \sum_\alpha \left( \frac{\overline{W}}{W_\alpha} - \frac{h_{s,\alpha}}{c_p T} \right) \left[ \dot{m}_\alpha'''  + \dot{m}_{b,\alpha}''' - \nabla \cdot (\mathbf{J_{d}}_\alpha) - \mathbf{u} \cdot \nabla (\rho Y_\alpha) \right] \label{eq:divth}
\end{eqnarray}
where $h_{s,\alpha}$ is the sensible enthalpy of species $\alpha$ and $h_s$ is the sensible enthalpy of the mixture, $\dot{q}''', \dot{q}_b'''$ are heat release rates due to combustion and particle gasification, $T$ is the local gas temperature, and $w$ is the local vertical velocity. The mixture specific heat at constant pressure and molecular weight are $c_p=\sum_{\alpha =1}^N{c_{p,\alpha} Y_\alpha}$ and $\overline{W}=\left(\sum_{\alpha =1}^N{Y_\alpha /W_\alpha} \right)^{-1}$ respectively. This thermodynamic divergence is derived from and acts as a proxy for the sensible enthalpy evolution equation~\cite{mcdermo_2014}. Equation~(\ref{eq:divth}) is derived by factoring the divergence from the sensible enthalpy equation and applying the ideal gas law.  Given a background pressure and a local mass density, the local temperature is computed from the ideal gas law as $T = \bar{p} \overline{W} / (\rho R) $.

The term $\dot{\mathbf{q}}''$ in Eq.~(\ref{eq:divth}) represents the conductive and diffusive heat fluxes discussed in Sec~\ref{sec:subgrid}. The net contribution from thermal radiation in the energy equation is defined by:
\begin{equation}
   -\nabla \cdot \dot{\mathbf{q}}_{\rm r}''(\bx) =
    \kappa(\bx) \, \left[ U(\bx) - 4 \pi \, I_{\rm b}(\bx) \right]  \quad ; \quad
    U(\bx) = \int_{4\pi} \, I(\bx,\bs') \, d\bs'  \label{simple_rte}
\end{equation}
where $\kappa(\bx)$ is the absorption coefficient, $I_b(\bx)$ is the source term, and $I(\bx,\bs)$ is the solution of the radiation transport equation (RTE) for a non-scattering gray gas:
\begin{equation}
   \bs \cdot \nabla I(\bx,\bs) = \kappa(\bx) \; \left[ I_{\rm b}(\bx) - I(\bx,\bs) \right] \label{bandRTE1}
\end{equation}
The source term, $I_{\rm b}$, requires special treatment because of the limited resolution of the underlying numerical grid in the vicinity of flames. In large scale fire simulations, grid cells are typically on the order of
tens of centimeters. Flame sheets cannot be resolved, meaning that the computed cell-average temperature can be significantly lower than temperatures one would expect to find in the reacting flame. Consequently, the
source term is approximated in grid cells where fuel and oxygen react. Elsewhere, the subgrid temperature field is homogeneous and the source term can be computed directly:
\begin{equation} \kappa I_{\rm b} = \left\{ \begin{array}{cll}
    \kappa \sigma T^4/\pi      & \hbox{Outside flame zone}, & \dot{q}'''=0  \\ [0.1in]
    C \kappa \sigma T^4/\pi  & \hbox{Inside flame zone}, & \dot{q}'''>0
    \end{array} \right.  \label{radapprox1}
\end{equation}
The constant $C$ is computed at each time step so that the volume integral of Eq.~(\ref{simple_rte}) over the entire flaming region is approximately equal to the volume integral of $\chi_{\rm r} \,\dot{q}'''$ over that same region. Here, $\chi_{\rm r}$ is an empirical estimate of the {\em global} fraction of that energy emitted as thermal radiation. Typically, a sooty fire radiates approximately one-third of the total combustion energy.

The radiation equation is solved using a technique similar to a finite volume method for convective transport, thus the name given to it is the Finite Volume Method (FVM). Using approximately 100 discrete angles which are updated over multiple time steps, the finite volume solver requires about 20\% of the total CPU time of a calculation, a modest cost given the complexity of radiation heat transfer.

Section~\ref{particle_model} discusses how Lagrangian particles are used to represent vegetation like leaves, pine needles and so on. These particles absorb and emit thermal radiation, and the underlying assumptions are described in that section.

\subsection{Subgrid Parameterizations} \label{sec:subgrid}

In the formal derivation of the LES equations (see, e.g, \cite{Pope:2000,FDS_Math_Guide}) the filtered nonlinear terms (advection terms, mean chemical source term, and the unresolved boundary fluxes) require subgrid-scale parameterizations.  In this section we will temporarily utilize the LES filter formalism as well as Cartesian tensor index notation to clarify the terms used in the governing equations.

\subsubsection{Subgrid Advection}

Unresolved turbulent eddies enhance the transport of mass, momentum, and energy.  This added transport is accounted for with residual stresses and scalar fluxes.  The subgrid-scale stress tensor is defined as
\begin{equation}
\tau_{ij}^{sgs} \equiv \bar{\rho}(\widetilde{u_i u_j} - \tilde{u}_i \tilde{u}_j)
\end{equation}
where an overline represents a volumetric filter and the tilde represents a Favre filter (see, e.g., \cite{Poinsot:TNC}).

In FDS the deviatoric part of the subgrid stress tensor is modeled using the gradient diffusion hypothesis.  Thus, the combined deviatoric stress may be written as follows:
\begin{equation}
\tau_{ij}^{dev} = \overline{\tau}_{ij} + \tau_{ij}^{sgs} - \frac{1}{3} \tau_{kk}^{sgs} = -2 (\mu + \mu_t) \left[ \tilde{S}_{ij} - \frac{1}{3} \tilde{S}_{kk} \right]
\end{equation}
where the symmetric rate-of-strain tensor is $S_{ij} \equiv \frac{1}{2}(\partial u_i/\partial x_j + \partial u_j/\partial x_i)$ and $\mu_t$ is an isotropic eddy viscosity.  Note that $\overline{\tau}_{ij}$ is already the deviatoric part of the molecular stress as the isotropic part is the pressure; $\mu$ is the molecular dynamic viscosity.

The eddy viscosity $\mu_t$ is an important quantity in the simulation as it affects all the turbulent transport and subgrid mixing time scales, which, as we will see below, are crucial to the combustion model.  In FDS, the eddy viscosity is modeled as \cite{Pope:2000,Deardorff:1980}
\begin{equation}
\mu_t = \bar{\rho} C_\nu \Delta \sqrt{k_{sgs}}
\end{equation}
with the constant taken as $C_\nu = 0.1$ (this value can be derived assuming production equals dissipation and a model Kolmogorov spectrum \cite{Pope:2000}; we also test the model against decaying isotropic turbulence \cite{CBC}).  In practice, the filter scale $\Delta$ is taken as the cube root of the cell volume.  Deardorff \cite{Deardorff:1980} solved a transport equation for the subgrid kinetic energy per unit mass, $k_{sgs}$, in order to account for subgrid buoyancy. In FDS, the density field is sufficiently resolved to account for buoyant plume accelerations, and therefore we find that an algebraic closure for $k_{sgs}$ is sufficient.  We utilize a scale-similarity idea, following the work of Bardina \cite{Bardina:1980}, and model the subgrid kinetic energy per unit mass as follows:
\begin{equation}
k_{sgs} = \frac{1}{2}(\tilde{u}_i - \hat{\tilde{u}}_i)(\tilde{u}_i - \hat{\tilde{u}}_i)
\end{equation}
where the hat represents a test filter at a scale $2\Delta$.  For the first Cartesian cell off the wall, where a test filter is not well defined, the eddy viscosity is taken from the WALE model \cite{Nicoud:1999}.

Subgrid advection of mass and heat are also modeled using gradient diffusion.  The turbulent diffusivities use constant turbulent Schmidt and Prandtl numbers, both set to 0.5; this simplification is justified based on high resolution simulations of fire plumes (see, e.g., \cite{Maragkos:2020}). For mass diffusion of species $\alpha$ in direction $i$ we have
\begin{equation}
J_{\alpha,i} = \bar{J}_{\alpha,i} + J_{\alpha,i}^{sgs} = -\bar{\rho}(D_\alpha + \frac{\mu_t}{\mathrm{Sc}_t}) \frac{\partial \tilde{Y}_\alpha}{\partial x_i}
\end{equation}
We use effective thermal conductivity and mass diffusivity in the heat flux, which may be written as
\begin{equation}
{\dot{q}''}_i = -(\bar{k} + \frac{\mu_t \bar{c}_p}{\mathrm{Pr}_t}) \frac{\partial \widetilde{T}}{\partial x_i} - \sum_\alpha \tilde{h}_{{\rm s},\alpha} ( \bar{\rho} D_\alpha + \frac{\mu_t}{\mathrm{Sc}_t} ) \frac{\partial \tilde{Y}_\alpha }{\partial x_i}
\end{equation}

\subsubsection{Filtered Chemical Source Term}

FDS treats combustion chemistry using a simplified approach.  We track only Fuel, Air, and Products, which we refer to as ``lumped species''.  Lumped species are groups of species that transport and react together and are always found in the same proportion.  For example, Air is a lumped species made up of 23 \% O$_2$ and 77 \% N$_2$ by mass.  For vegetation, a surrogate hydrocarbon Fuel is usually assumed.  For example, cellulose may be assumed to decompose to a volatile fuel gas with the formula $\C_{6} \H_{10} \O_{5}$.  The Products are defined stoichiometrically, often including prescribed yields of soot (for smoke) and carbon monoxide that may be obtained from empirical relationships for the specific fuel or from other sources (e.g., \cite{SFPE:Mulholland}).  Then, usually, a single reaction is tracked\footnote{More complicated reaction schemes are possible in the code, but for this work a single reaction is sufficient.}:
\begin{equation}
\mathrm{Fuel} + s \, \mathrm{Air} \rightarrow (1+s) \,\mathrm{Products}
\end{equation}
where $s$ is the mass stoichiometric coefficient.

In LES of turbulent combustion the filtered chemical source term, $\overline{\dot{m}_\alpha'''}$, is unclosed because the reaction kinetics are nonlinear functions of species concentration and temperature.  However, since the heat releasing reactions in a fire are extremely fast compared to the time scales for turbulent mixing, it is common practice in fire dynamics simulations to utilize the eddy dissipation model of Magnussen and Hjertager \cite{Magnussen:1} (see also \cite{Poinsot:TNC}).  The mass production term for the Fuel species is written as
\begin{equation}
\overline{\dot{m}_{F}'''} = -\bar{\rho} \,\frac{\min(\tilde{Y}_F,\tilde{Y}_A/s)}{\tau_{mix}}
\end{equation}
This formula states that the fuel is consumed at a rate proportional to the local limiting reactant concentration and inversely proportional to the local mixing time scale.  This is the so-called ``mixed is burnt'' approximation.  The mixing time scale is given by $\tau_{mix} = \min(\tau_d, \tau_u,\tau_g)$ \cite{McDermott:2011}, where the time scales for diffusion, turbulent advection, and gravitational acceleration are, respectively,
\begin{align}
\tau_d &= \Delta^2/D_F \\
\tau_u &= C_u \Delta / \sqrt{(2/3)k_{sgs}} \\
\tau_g &= \sqrt{2\Delta/g}
\end{align}
Here, $D_F$ is the binary diffusivity of the Fuel species in Air, $k_{sgs}$ is the modeled subgrid kinetic energy per unit mass discussed above, and $g$ is the gravitational acceleration. The constant $C_u=0.4$ is determined by matching flame height correlations across a wide range of fire Froude numbers \cite{FDS_Validation_Guide}.

The remaining mass production rates are found from stoichiometry.  The local heat release rate, $\dot{q}'''$, an important term in Eq.~(\ref{eq:divth}), is then determined from
\begin{equation}
\label{eq:qdotppp}
\dot{q}''' = - \sum_\alpha \overline{\dot{m}_\alpha'''} \Delta h_{f,\alpha}
\end{equation}
where $\Delta h_{f,\alpha}$ are the heats of formation for each lumped species $\alpha$.

\paragraph{Flame Extinction} One of the advantages of a physics-based tool like FDS is the possibility of building into the model the effects of defensive actions on the fire spread rate.  FDS uses simple empirical rules to predict local extinction within a gas phase grid cell based on resolved species concentrations and the mean cell temperature. The rules are based on the concept of a critical flame temperature. The basic theory behind the critical flame temperature is described in \cite{SFPE:Beyler}.  In brief, if the local heat release given by Eq.~(\ref{eq:qdotppp}) is not sufficient to raise the cell temperature above the CFT, then $\overline{\dot{m}_{F}'''}$ is set to zero for that cell for that time step.  Note that the flame extinction model does not directly affect surface cooling or smothering of the vegetative fuel, which may happen with certain fire suppression activities.  However, such tactics can be accounted for through modifications to the solid fuel thermal degradation model discussed below.

\subsubsection{Unresolved Boundary Fluxes}

In this section we discuss the wall functions used to close the mass, momentum, and energy fluxes on solid boundaries.  Mass transfer of moisture and fuel from the surface is discussed in more detail below in Sec.~\ref{sec:firespread} where the various flames spread models are compared (which is really the main point of this paper).

The scaled streamwise velocity is denoted $u^+ = u/u_\tau$, where the friction velocity is $u_\tau \equiv \sqrt{\tau_w/\rho}$ and $\tau_w$ is the wall stress needed to close the momentum equation, Eq.~(\ref{eqn:momtran}).  For rough walls FDS employs the log law presented in \cite{Pope:2000},
\begin{equation}
\label{eqn_roughwallloglaw}
u^+ = \frac{1}{\kappa} \ln \left(\frac{y}{s}\right) + \tilde{B}(s^+)
\end{equation}
where $\kappa = 0.41$ is the von K\'arm\'an constant, $s^+ = s/\delta_\nu$ is the roughness length in viscous units, $s$ is the dimensional ``sand grain'' roughness, and $\delta_\nu = \mu/(\rho u_\tau)$ is the viscous length scale. The distance to the wall, $y$, is taken as $\delta y/2$ for the first off-wall grid cell.  The parameter $\tilde{B}$ varies with $s^+$ but attains a constant value $B_2=8.5$ in the fully rough limit.  In FDS, we implement $\tilde{B}$ as the following piece-wise function:
\begin{equation}
\tilde{B} = \left\{ \begin{array}{lll} B + (1/\kappa)\ln(s^+) & \mbox{for} & s^+ < 5.83 \\
\tilde{B}_{\rm max} & \mbox{for} & 5.83 \le s^+ < 30.0 \\
B_2 & \mbox{for} & s^+ \ge 30.0 \end{array} \right.
\end{equation}
where $\tilde{B}_{\rm max} = 9.5$.

In the fully rough limit the sand grain roughness may be equated to the aerodynamic roughness, $z_0$, typically employed in atmospheric codes for a neutral boundary layer,
\begin{equation}
s = z_0 e^{8.5 \kappa} \approx 32.6 \, z_0
\end{equation}
It is important to appreciate that FDS was originally designed to handle compartment fires and must consider flows with smooth walls such as HVAC ducts.  Further, FDS typically resolves the thermal motions that would necessitate the use of Monin-Obukhov stability corrections to the boundary profile.

The unresolved heat flux at the surface, $\dot{q}_w''$, is determined using an engineering approach,
\begin{equation}
\dot{q}_w'' = h (T_g - T_w) 
\end{equation}
where $h$ is the convective heat transfer coefficient, $T_w$ is the local temperature of the surface, and $T_g$ is the temperature of the first gas phase cell off the wall.  Empirical natural/forced convection correlations are used to determine $h$:
\begin{equation}
h = \max \left[C\, |T_g-T_w|^{1/3} \; , \;
\frac{k}{L} \, \mathrm{Nu} \; , \;
\frac{k}{\delta n/2} \right]
\label{eq:qconv}
\end{equation}
where $C$ is a empirical coefficient for natural convection (1.52 for a horizontal plate and 1.31 for a vertical plane or cylinder)~\cite{Holman:1}, $L$ is a characteristic length related to the size of the physical obstruction, and $k$ is the thermal conductivity of the gas. The forced convection Nusselt number (Nu) depends on the geometry and flow characteristics \cite{Holman:1,Incropera:1}.  Further details are provided in \cite{FDS_Math_Guide}.
